
---
title: "Reproducible Research: Peer Assessment 1"
author: "Bill Schroeder"
output: 
  html_document:
    keep_md: true
---

```{r setoptions, echo=FALSE, cache=FALSE}
options(scipen = 999, digits = 2)

```

## Loading and preprocessing the data
  
  

* Check for and install required libraries

```{r checkForLubridateChunk}

# check for required packages
if("lubridate" %in% rownames(installed.packages() ) == FALSE){
  install.packages("lubridate")}
library(lubridate)

```

* Load input file

```{r loadDataChunk}

# read in the data
if(!file.exists("activity.csv")){
  unzip("activity.zip")
}
data <- read.csv( "activity.csv", header=TRUE, sep=",")

```


* Process/transform the data into a format suitable for your analysis

```{r convertDatesChunk}

# convert the dates (y/m/d) from factors to dates
data$date <- ymd(as.character(data$date))

```


## What is mean total number of steps taken per day? 
  
  

```{r, fig.height=4, echo=FALSE}

# calculate the sum of steps per date
sumStepsPerDay <- tapply (data$steps, data$date , sum)

# create a histogram 
hist( sumStepsPerDay, xlab="Steps/Day", ylab="Frequency", main="Histogram Total Steps per Day" )

```


* Calculate the mean and median total number of steps taken per day



```{r}

sumStepsPerDay <- tapply (data$steps, data$date , sum)
meanTotalSteps <- mean( sumStepsPerDay, na.rm = TRUE )
medianTotalSteps <- median( sumStepsPerDay, na.rm = TRUE )

```

* The mean of the total number of steps taken per day is **`r meanTotalSteps`** 

* The median of the total number of steps taken per day is **`r medianTotalSteps`**


## What is the average daily activity pattern?
  
  
* Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}

# calculate the mean number of steps for each interval
meanStepsPerInterval <- tapply (data$steps, data$interval, mean, na.rm=TRUE)

# find the maximum mean value across all intervals
maxSteps <- max(meanStepsPerInterval) 

# find the interval that contains the maximum mean value
maxStepsInterval <- match( maxSteps, meanStepsPerInterval )

# create a time series plot of steps for all intervals 
plot(meanStepsPerInterval , type="l", xlab="interval", ylab="ave/steps")
abline( v=maxStepsInterval, lty=2, col="red", lwd="5")
text(maxStepsInterval,10,maxStepsInterval) 

```


* Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}

# find the maximum mean value across all intervals
maxSteps <- max(meanStepsPerInterval) 

# find the interval that contains the maximum mean value
maxStepsInterval <- match( maxSteps, meanStepsPerInterval )

```

* Report the 5-minute interval that, on average, contains the maximum number of steps
    - The interval **`r maxStepsInterval`** contains the ave maximimum number of steps, which is **`r maxSteps`**  
    

## Imputing missing values
  
  
* Calculate and report the total number of missing values in the dataset  
  
```{r}
# function to calculate total number of NAs
numNAs <- function( data ) {
	isNA <- is.na(data$steps)
	return (sum(isNA))
}

```

* Total number of NAs = **`r numNAs( data )`** Which is **`r mean(is.na(data$steps))*100`%**


* Devise a strategy for filling in all of the missing values in the dataset. Strategy is to replace NAs with ave/day
      + **Fill all mission values with the average steps per day**

* Create a new dataset that is equal to the original dataset but with the missing data filled in

```{r}

# clean the original data - replace NAs with average for day
cleansedData <- data
cleansedData$steps[ is.na(cleansedData$steps)] <- mean(cleansedData$steps, rm.na=TRUE)

```


* Make a histogram of the total number of steps taken each day

```{r echo=FALSE}

# calculate the sum of steps per date
sumStepsPerDayCleansed <- tapply (cleansedData$steps, cleansedData$date , sum)

# create a histogram of the sum of steps per day
hist(sumStepsPerDayCleansed, xlab="Steps/Day", ylab="Frequency", main="Total steps per day - cleansed data set")

```


* Calculate and report the mean and median total number of steps taken per day


```{r}

# calculate the mean across all dates
mean(sumStepsPerDayCleansed, na.rm=TRUE)

# calculate median across all dates
median(sumStepsPerDayCleansed, na.rm=TRUE)

```


* Do these values differ from the estimates from the first part of the assignment? 
* What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}

meanDif <- mean(cleansedData$steps, na.rm=TRUE) -  mean( data$steps, na.rm = TRUE ) 
medianDif <- median(cleansedData$steps, na.rm=TRUE) -  median( data$steps, na.rm = TRUE )

if(meanDif != 0){
  print(sprintf("Cleaning the data DID impact the mean by %s ", meanDif) )
} else {
  print("Cleaning the data did NOT impact the mean")
}

if(medianDif != 0){
  print(sprintf("Cleaning the data DID impact the median by %s ", medianDif) )
} else {
  print("Cleaning the data did NOT impact the median")
}

```


## Are there differences in activity patterns between weekdays and weekends?

* Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.*

```{r}
# check if package "chron" is installed
if("chron" %in% rownames(installed.packages() ) == FALSE){
  install.packages("chron")
}
library("chron")

# function returns if date is a weekend or weekday
dayOfWeekFactor <- function( date ){
  if( TRUE == factor(is.weekend(date) )){ 
    "weekend"
      } else {
        "weekday"
    }
}

# Create a new factor variable in the dataset with two levels# "weekday" and "weekend" 
cleansedData$DoW <- as.factor(sapply(cleansedData$date, dayOfWeekFactor))

```

* Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 


```{r echo=FALSE}

# create time series of intervals and steps on weekdays
weekdayData <- cleansedData[cleansedData$DoW == "weekday",]
weekdayData <- tapply (weekdayData$steps, weekdayData$interval, mean, na.rm=TRUE)

# create time series of intervals and steps on weekend
weekendData <- cleansedData[cleansedData$DoW == "weekend",]
weekendData <- tapply (weekendData$steps, weekendData$interval, mean, na.rm=TRUE)


# create a timeseris panel plot of ave steps taken on weekend vs weekdays
par(mfrow=c(2,1))
par(mai = c(0, 1, 1, 0))
plot(weekendData, type="l", xaxt="n", ylab="ave steps")
legend('topright',"WeekEND")
title("Week day/end ave steps per interval")
par(mai = c(1, 1, 0, 0))
plot(weekdayData, type="l", xlab="interval", ylab = "ave steps")
legend('topright',"WeekDAY")


```

  